---
title: "Proj_Code_Analysis"
output:
  pdf_document: default
  html_document: default
date: "2025-03-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load("/Users/ellathomasson/Documents/STAT 4996/NSDUH_2023.Rdata")
library(dplyr)
library(ggplot2)
library(kableExtra)
```

```{r}
povertyDepData<-puf2023_102124[,c("AMDELT",
                                  "POVERTY3", 
                                  "NEWRACE2",
                                  "IRHHSIZ2",
                                  "MOVSINPYR2",
                                  "MEDICARE",
                                  "CAIDCHIP",
                                  "CHAMPUS",
                                  "PRVHLTIN",
                                  "GRPHLTIN",
                                  "IREDUHIGHST2",
                                  "WRKSTATWK2", 
                                  "CIGREC", 
                                  "HLTINALC", 
                                  "HLTINDRG", 
                                  "CIGREC"
)]

names(povertyDepData)<-c("lifetimeDepression",
                         "povertyThreshold",
                         "race",
                         "householdSize",
                         "timesMovedPastYear",
                         "medicare",
                         "medicaid_chip",
                         "govtHealthcare",
                         "privateHealthcare",
                         "employer_unionHealthcare",
                         "highestEducationCompleted",
                         "workStatusPastWeek",
                         "LastSmoke", 
                         "AlcHealthCare",
                         "DrgHealthCare", 
                         "timeSinceLastCigarette")
```


# analysis


```{r}
sum(is.na(povertyDepData$lifetimeDepression)) / nrow(povertyDepData)
```


## cleaning 

```{r}
povertyDepData %>%
  group_by(householdSize) %>%
  summarize(
    count = n()
  )
```


```{r}
povertyDepData_cleaned <- povertyDepData[!is.na(povertyDepData$lifetimeDepression), ]
povertyDepData_cleaned <- povertyDepData_cleaned[!is.na(povertyDepData_cleaned$householdSize), ]

povertyDepData_cleaned = povertyDepData_cleaned[povertyDepData_cleaned$timesMovedPastYear < 4,]
povertyDepData_cleaned = povertyDepData_cleaned[povertyDepData_cleaned$medicare < 3,]

povertyDepData_cleaned = povertyDepData_cleaned[,c("lifetimeDepression","race","timesMovedPastYear",
                                                   "highestEducationCompleted","medicare","timeSinceLastCigarette","householdSize")]



povertyDepData_cleaned$timeSinceLastCigarette <- as.numeric(povertyDepData_cleaned$timeSinceLastCigarette)


povertyDepData_cleaned <- povertyDepData_cleaned %>%
  mutate(
    timeSinceLastCigarette = case_when(
      timeSinceLastCigarette == 1 ~ "Past 30 Days",
      timeSinceLastCigarette %in% c(2, 8, 19) ~ "Past Year",
      timeSinceLastCigarette %in% c(3, 14, 29, 39) ~ "Past 3 Years", 
      timeSinceLastCigarette%in% c(4,9) ~ "More than 3 Years",
      timeSinceLastCigarette == 91 ~ "Never Used",
      TRUE ~ as.character(timeSinceLastCigarette)  
    )
  )

povertyDepData_cleaned <- povertyDepData_cleaned %>%
  mutate(
    race = case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black/Afr Am",
      race == 3 ~ "Native Am/AK Native", 
      race == 4 ~ "Native HI/Other Pac Isl",
      race == 5 ~ "Asian",
      race == 6 ~ "more than one race",
      race == 7 ~ "Hispanic",
      TRUE ~ as.character(race)  # Keep original value (converted to character)
    )
  )

```

Current Explan Variables: "lifetimeDepression","race","timesMovedPastYear","highestEducationCompleted","medicare","timeSinceLastCigarette","householdSize"


```{r}
(povertyDepData_cleaned) %>%
  group_by(timeSinceLastCigarette) %>%
  summarise(
    n()
  )
```


## train-test split 

```{r}
povertyDepData_cleaned$lifetimeDepression <- ifelse(povertyDepData_cleaned$lifetimeDepression == 1, 1, 0) # 1 is "has had depression"
povertyDepData_cleaned$medicare <- ifelse(povertyDepData_cleaned$medicare == 1, 1, 0) # 1 is "covered by medicare" 

povertyDepData_cleaned$lifetimeDepression = as.factor(povertyDepData_cleaned$lifetimeDepression)
povertyDepData_cleaned$medicare = as.factor(povertyDepData_cleaned$medicare)

povertyDepData_cleaned

```


```{r}

# One-hot encode the dataset 
encoded_data <- data.frame(
  model.matrix(~ . - lifetimeDepression - 1, data = povertyDepData_cleaned)
)

# Add back the response variable
encoded_data$lifetimeDepression <- povertyDepData_cleaned$lifetimeDepression

# split into train and test sets
sample_size <- floor(0.7 * nrow(encoded_data))
train_indices <- sample(seq_len(nrow(encoded_data)), sample_size)
train_data <- encoded_data[train_indices, ]
test_data <- encoded_data[-train_indices, ]


```


# building the model

```{r}

logistic_model <- glm(lifetimeDepression ~ ., data = train_data, family = "binomial")

```


```{r}
stepwise_logistic_model<-stepAIC(logistic_model, direction = "both")
```


## refitting the model based on the stepwise regression

```{r}
logistic_model_step <- glm(
  lifetimeDepression ~ raceAsian + raceBlack.Afr.Am + raceHispanic + 
    racemore.than.one.race + raceNative.Am.AK.Native + raceNative.HI.Other.Pac.Isl + 
    timesMovedPastYear + highestEducationCompleted + medicare1 + 
    timeSinceLastCigaretteNever.Used + timeSinceLastCigarettePast.3.Years + 
    timeSinceLastCigarettePast.30.Days + timeSinceLastCigarettePast.Year + 
    householdSize,  
  data = train_data, 
  family = binomial(link = "logit")
)
```



```{r}
predictions <- predict(logistic_model_step, newdata = test_data, type = "response")
predicted_class <- ifelse(predictions > 0.5, 1, 0)

hist(predictions)

mean(predictions)
```


```{r}
confusion_matrix <- table(
  Actual = test_data$lifetimeDepression, 
  Predicted = predicted_class
)
print("Confusion Matrix:")
print(confusion_matrix)


# False Positive Rate
FPR <- confusion_matrix[1,2] / (confusion_matrix[1,2] + confusion_matrix[1,1])
print(paste("False Positive Rate:", round(FPR, 4)))

# False Negative Rate
FNR <- confusion_matrix[2,1] / (confusion_matrix[2,1] + confusion_matrix[2,2])
print(paste("False Negative Rate:", round(FNR, 4)))

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 4)))
```

```{r}
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 4))) 
```


The fact that the classes are unequal in size seems to be causing issues. This may need to be addressed sooner rather than later. 

Also, may need to try more complex model? 

```{r}
library(car)
vif(logistic_model_step)
```


